A Mechanised Library for Generic (Co)Programming

### 2023-12-31 paper/paper.tex
Draft introduction.

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While both \haskell{partition} and the \haskell{quicksort} program using it are
well-defined functions a proof assistant such as Agda or Coq cannot accept such
definitions.

The argument can be given on pen-and-paper by showing the \haskell{partition} is
a function that always returns two sublists which are smaller than the one we
started with which means the repeated application of \haskell{partition} will
eventually lead to termination.

The quicksort algorithm is a \emph{divide-and-conquer} algorithm. Such
algorithms are structured in three parts: first, the input is broken down in
smaller sub-problems by means of a \emph{well-founded coalgebra}.Secondly, these
sub-problems are computed recursively and, finally, the results of the recursive
call are put together by a an algebra.

Practically, every program can be turned into an hylomorphism by means of the
conjugate rule~\cite{HinzeWG15}. This can be paraphrased by saying
\begin{center}
  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
\end{center}

This has been a useful for structuring functional programs in a way that it
guarantees termination (or productivity) even in programming languages with
general recursion~\cite{DanielssonHJG06}.

With this work we intend to begin a study of recursion scheme beyond the world
of functional programming to the one of dependently typed languages such as a
the ones employed in proof assistants like Agda or Coq.

>>>>

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While \haskell{quicksort} is a well-defined mathematical function, a proof
assistant such as Agda or Coq is not able to accept such definitions. The reason
is that the function \emph{destructuring} the input, that is the
\haskell{partition} function is not simply destructuring the list returning the
head and the tail of it, but it returns two sublists with the head as a pivot.

While the syntactic checker of a proof assistant cannot understand that these
two sublists are eventually going to be smaller than the original input we can
can give a pen-and-paper by showing the \haskell{partition} is a
\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
type $A$. Here the term ``well-founded'' refers to the fact that the infinite
iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
in $A$.

More generally, this is an instance of a \emph{divide-and-conquer} algorithm.
Such algorithms are structured in three parts: first, the input is broken down
in smaller sub-problems by means of a well-founded coalgebra. Secondly,
these sub-problems are computed recursively and, finally, the results of the
recursive call are put together by a an algebra. This particular way of
doing recursion has been named in the functional programming literature as a hylomorphism~\cite{MeijerFP91, HuIT96}.

Many years later Hinze et al.~\cite{HinzeWG15} showed that every recursive
program can be turned into an hylomorphism by means of the conjugate rule. The
essence of this result is that every complex recursion scheme can be explained by its associated \emph{adjunction} which establishes a way of reducing the proof obligation
for the complex recursion scheme into the proof obligation for a basic hylomorphism.
This can be paraphrased by saying
\begin{center}
  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
\end{center}

Since then this result has had huge impact in the functional programming
community for structuring functional programs in a way that it guarantees
termination (or productivity) in programming languages with general
recursion~\cite{DanielssonHJG06}.

There is a tension:
\begin{itemize}
\item On one hand we can allow general recursion and the employ recursion schemes to help the programmer structuring programs so that they terminate
  \item on the other hand we could forbid non-termination, but at this point there is no algorithm to tell termination and non-termination apart.
\end{itemize}

The first option, helps the programmer but does not guarantee termination, since the programmer can still feed to the recursion scheme a functor that does not have a fixed-point
or that has an algebra that does not terminate.

The second option forbids non-termination, but does not allow all terminating programs.

With this work we intend to begin a study of recursion scheme beyond the world
of functional programming to the one of dependent type theories such as a the
ones employed in proof assistants like Agda or Coq where non-termination is not
included in the language. For type theories it has been a long-standing
challenge to come up with a sound algorithm which rejects all programs that do
not terminate (soundness) and accepts as many well-defined programs as possible
(completeness). Our aim is to provide an additional tool to the every-day
programmer to be able to extend the range of recursive programs we can write.
